%% How to compile:
%% $ pdflatex report.tex
%% $ bibtex report.aux
%% $ pdflatex report.tex

%% Guidelines on the ACM Latex package:
%% https://authors.acm.org/proceedings/production-information/preparing-your-article-with-latex


%% format options: manuscript, acmsmall, acmlarge, acmtog, sigconf, siggraph, sigplan, sigchi
%% used formats: manuscript, sigplan
\documentclass[
  format=manuscript,
  screen=true,
  review=false,
  nonacm=true,
  timestamp=true,
  balance=false]{acmart}
\setcopyright{rightsretained}

\author{Alexander Christensen}
\title{C++ Scoped Enum Enhancements}
\email{<alex\_c007@hotmail.com>}

%% \citestyle{acmauthoryear}


%% ===================== %%
%% === CODE LISTINGS === %%
%% ===================== %%
\usepackage{listings}
\usepackage{color}

\definecolor{backgroundblue}{rgb}{0.93, 0.93, 1.0}
\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}

\definecolor{LightCyan}{rgb}{0.88,1,1}
\definecolor{bluekeywords}{rgb}{0.13,0.13,1}
\definecolor{greencomments}{rgb}{0,0.5,0}
\definecolor{turqusnumbers}{rgb}{0.17,0.57,0.69}
\definecolor{redstrings}{rgb}{0.5,0,0}
\definecolor{red}{rgb}{0.5,0.0,0.0}
\definecolor{blue}{rgb}{0.0,0.5,1.0}
\definecolor{green}{rgb}{0.0,0.5,0.0}


% basic settings, can be overrided
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  columns=fullflexible,
  escapeinside={\%},
  frame=single,
  backgroundcolor=\color{backgroundblue},
  showspaces=false,
  keepspaces=true,
  showstringspaces=false,
  showtabs=false,
  numbers=left,
  aboveskip=-3pt,
  sensitive=true
}


% nice horizontal line "-------- <text> --------", used above a listing
\def\headline#1{\begin{minipage}{36em}\vspace{4mm}\hrulefill\quad\lower.3em\hbox{#1}\quad\hrulefill\end{minipage}}

% horizontal line, used below a listing
\newcommand{\lstline}{\vspace{-3mm}\hrulefill\vspace{2mm}\newline}

% fancy insert listing command
% example of usage: `\customlisting{FSharp}{Some Function}{<file.fsx>}`
\newcommand{\customlisting}[3]{\lstinputlisting[language=#1,name=#2]{#3}\lstline}

\newcommand{\beforecodespace}{\vspace{4mm}}


% Insert the code, e.g. from a file like this:
% \lstinputlisting[language=<language>]{<input_file>}
% \lstinputlisting[language=<language>, firstline=X, lastline=Y]{<input_file>}

% Or directly write the code in the .tex document:
% \begin{lstlisting}[language=<language>]

\lstdefinelanguage{Cpp}{
  morekeywords={float, int, double, uint, bool, if, for, else, void, class,
                struct, private, protected, public, enum, const, char, case,
                default, return, switch, template, constexpr, noexcept, auto,
                typename, namespace, concept, static, unsigned, consteval,
                long, using},
  keywordstyle=\color{bluekeywords},
  morecomment=[l][\color{greencomments}]{///},
  morecomment=[l][\color{greencomments}]{//},
  morecomment=[l][\color{redstrings}]{\#},
  morecomment=[s][\color{greencomments}]{{/*}{*/}},
  morestring=[b]",
  stringstyle=\color{redstrings},
  %title={\headline{\small C++ - \textit{\lstname}}}
}


%% =========================== %%
%% === ADDITIONAL PACKAGES === %%
%% =========================== %%
% Load with some options, i.e. \usepackage[colorlinks=true,linkcolor=blue]{hyperref} or blank
% \usepackage{hyperref}
%\usepackage[colorlinks=true,linkcolor=blue, linktocpage]{hyperref}
\usepackage{stackengine}
\newlength\llength
\llength=\linewidth\relax


%% ================ %%
%% === DOCUMENT === %%
%% ================ %%
\begin{document}

%% \abstract{
%% This is a very abstract abstract.
%% }

\maketitle
\tableofcontents

\section{Introduction}

This proposal is the first revision of "C++ Scoped Enum Proposal", which was circulated
in the C++ proposals email list during the winter of 2021/2022. The goal of writing
this proposal is to gain feedback and insights, which may then be used for further
revisions.

\noindent
\textbf{Note:} This proposal targets only \underline{scoped enums}, and does not provide any
effort to address \textit{unscoped} enums.

\vspace{2mm}\noindent
Given my limited experience (ie. none) in writing C++ proposals, I know what I
\underline{would like} the proposal to address and introduce to the core language,
but not necessarily \underline{how it should be done}. The suggested notations
provided are, thus, merely suggestions - and not concrete proposals for a concious
choice of syntax.


\section{Motivation and Scope}

The initial motivation for this proposal was the lack of a good way in the standard
library to convert an enum value to a proper string representation. Very often, I
have found a need to log an enum for various purposes, and every time I create a new
enum type I have to write such a function again. An example:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
enum class GraphicsApiType { none, opengl, vulkan };

const char* get_api_type_string(GraphicsApiType apiType) {
    switch (apiType) {
    case GraphicsApiType::none:   return "none";
    case GraphicsApiType::opengl: return "opengl";
    case GraphicsApiType::vulkan: return "vulkan";
    default: return "<unrecognized>";
    }
}
\end{lstlisting}

\noindent
This is cumbersome to maintain, for every time a value is added or modified inside
the enum type, this other function has to be modified as well. But besides this,
a potential \textit{run-time} error may be introduced when an invalid enum is
provided.

A session of browsing around sites such as StackOverflow revealed sometimes quite
vividly imaginary answers for how to circumvent this limitation in the language:

\begin{enumerate}
\item \url{https://stackoverflow.com/questions/11421432/how-can-i-output-the-value-of-an-enum-class-in-c11}
\item \url{https://stackoverflow.com/questions/1390703/enumerate-over-an-enum-in-c}
\item \url{https://stackoverflow.com/questions/6281461/enum-to-string-c}
\item \url{https://stackoverflow.com/questions/201593/is-there-a-simple-way-to-convert-c-enum-to-string}
\end{enumerate}

\noindent
Quite intuitively, two key observations were made:

\begin{itemize}
\item This proposal introduces \underline{changes to the core language}.
\item While fixing enum to string functionality, there's a lot more that we could do at the same time.
\end{itemize}

\subsection{Scope for this proposal}

Only \textit{scoped enums} are extended with this proposal, because they are fully
type safe, meaning that we can theoretically perform all necessary computations needed
for this proposal \textit{compile-time}, with no risk of throwing exceptions or using
any other runtime mechanism.

This, at least was the aim. Added is one particular \textit{exception} to this,
where we try to convert to a scoped enum from its underlying type. This may not
neccessarily be possible to achieve compile-time, and will be described in detail.


\section{Impact on the Standard}

This proposal requires some features that as of this writing cannot be covered by
writing a library. These features \textit{might} possibly be implemented by the
compilers instead of modifying or extending the C++ standard. This is preferred.


\section{Proposed features}

\subsection{Concept for "is-an-enum"}

The basis for this proposal is a concept for restricting correct type substitution
on templated types. This can be implemented fully as a \textit{library feature},
using C++23:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
namespace std {
    template<typename T>
    concept scoped_enum = std::is_scoped_enum_v<T>;
}

template<std::scoped_enum T>
class foo { int bar; };

// Try it out:
foo<GraphicsApiType> f1;  // <-- success!
foo<unsigned long> f2;    // <-- fails to compile!
\end{lstlisting}


\subsection{Convert scoped enum to string}

Two ways of obtaining a string is proposed: A simple one which contains only the
name of the enum value, and a fully-qualified which \textit{additionally} contains
namespaces, classes, and other scopes:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
namespace std {
    template<std::scoped_enum T>
    constexpr std::string_view enum_str(T t) { /* compiler implements this */ }
}

std::cout << std::enum_str<GraphicsApiType::opengl>() << std::endl;      // "opengl"
std::cout << std::enum_str_full<GraphicsApiType::opengl>() << std::endl; // "GraphicsApiType::opengl"
\end{lstlisting}

\noindent
The return type itself gives us a number of possibilites, but two are probably worthy
of consideration: \texttt{const char*} and \texttt{std::string\_view}. It is probably
also worth considering what notation is \textit{optimal}. In the example above, we
have a templated function which return a \texttt{constexpr} string representation.

\vspace{2mm}\noindent
\textbf{Note:} I am unsure how best to write this.


\subsection{Obtain the underlying type}

In current C++, when needing to compare an enum value to its underlying type,
a relatively verbose expression is needed which involves a "\texttt{static\_cast}"
and possible also an "\texttt{std::underlying\_type}". This has been somewhat
mitigated by the introduction of \texttt{std::to\_underlying}, available with C++23,
but this proposal adds a new template meta-function which does almost the same
thing:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
template<scoped_enum E>
struct enum_type {
    using type = std::underlying_type<E>::type;
};

template<scoped_enum E>
using enum_type_t = std::enum_type<E>::type;

// Try it out:
enum class Color : int { red, green, blue, yellow };
std::enum_type_t<Color> yellow = Color::yellow;
static_assert(std::is_same_v<std::enum_type_t<Color>, int>);
\end{lstlisting}

The benefit of this meta-function, as opposed to \texttt{std::to\_underlying}, is
that this meta-function does not work for unscoped enumerations. This is because
\texttt{std::to\_underlying} uses \texttt{std::underlying\_type}, while this new
meta-function uses the newly proposed concept \texttt{std::scoped\_enum}.
This creates a more accentuated difference between scoped and unscoped enums, and
thereby encourages the use of scoped enums for enhanced type safety and scope
safety.


\subsection{Get the underlying value}

Similar to how getting an underlying type can be tedious, this proposal also
addresses the need for obtaining the underlying value of a scoped enum value:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
template<scoped_enum Enum>
constexpr std::enum_type_t<Enum> enum_value(Enum e) noexcept
{
    return static_cast<std::enum_type_t<Enum>>(e);
}
\end{lstlisting}


\subsection{Safe enum conversion (language feature)}

Sometimes, a conversion from an underlying type to an enum value may be approved
without being valid. And there exists no compile-time feature to check for this.
An example of how it may go wrong:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
enum class MyScopedEnum : int
{
    value_0 = 0, value_1 = 1, value_2 = 2, value_3 = 3, value_43 = 43, value_57 = 57
};

constexpr MyScopedEnum test = static_cast<MyScopedEnum>(56);
\end{lstlisting}

\noindent
Even if we explicitly denote our variable \texttt{constexpr}, the compiler is not
able to enforce correct behaviour, and so we obtain undefined behaviour. The topic
is discussed on StackOverflow:

\noindent
\url{https://stackoverflow.com/questions/17811179/safe-way-to-convert-an-integer-in-an-enum}

\vspace{3mm}\noindent
This proposal thus defines a template meta-function which fails at
\underline{compile-time} if the conversion is invalid. It might be possible to also
define a \textit{run-time} mechanism which throws an exception, but this revision
does not discuss that topic further (for now). Proposed syntax:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
template<std::scoped_enum T>
constexpr bool is_enum_convertible(std::enum_type_t<T> t) {
    return ...; // <-- compiler implements this
};

template<std::scoped_enum T>
consteval T enum_cast(std::enum_type_t<T> t) {
    static_assert(std::is_enum_convertible<T>(t));
    return static_cast<T>(t);
}
\end{lstlisting}

\noindent
\textbf{NOTE:} I am unsure of the best naming here, because the word "cast" does
not strictly enforce either compile-time or runtime behaviour. One might argue
that \texttt{enum\_cast} should be completely runtime and throw an exception,
and this proposal might be rewritten if such an opinion is present.


\subsection{IMPROVED Safe enum conversion (language feature)}

A \underline{alternative} to the \texttt{enum\_cast} suggested previously, we
could also extent \texttt{static\_cast} to correctly enforce compile-time check
of underlying type to enum conversion, and then instead reserve \texttt{enum\_cast}
for purely runtime purposes. This suggested, alternative suggestion is shown below:


%% NOTE: Borrowed from:
%% https://tex.stackexchange.com/questions/168741/showing-two-listings-in-a-table-side-by-side
\setbox0=\hbox{%
\begin{minipage}{0.45\linewidth}
\begin{center}
Current C++, \texttt{static\_cast} is unsafe.\vspace{2mm}
\end{center}
\begin{lstlisting}[language=Cpp, basicstyle={\footnotesize\ttfamily}]
enum class MyScopedEnum : int
{
    val_0 = 0, val_1 = 1, val_2 = 2
};

// FAIL: Compiler thinks this is okay
MyScopedEnum e = static_cast<MyScopedEnum>(4);
\end{lstlisting}
\end{minipage}
}
\savestack{\listingA}{\box0}

\setbox0=\hbox{%
\begin{minipage}{0.45\linewidth}
\begin{center}
Proposed restriction to \texttt{static\_cast}.\vspace{2mm}
\end{center}
\begin{lstlisting}[language=Cpp, basicstyle={\footnotesize\ttfamily}]
enum class MyScopedEnum : int
{
    val_0 = 0, val_1 = 1, val_2 = 2
};

// GOOD: Compiler throws an error
MyScopedEnum e = static_cast<MyScopedEnum>(4);
\end{lstlisting}
\end{minipage}
}
\savestack{\listingB}{\box0}

\begin{center}
\begin{tabular}{ccc}
\listingA & \hspace{5mm} & \listingB \\
\end{tabular}
\end{center}

\noindent
\textbf{NOTE:} This might be the best option. I will let feedback guide this
decision.


\subsection{Iterating through a scoped enum (language feature)}

Another interesting scenario is if an enum contains "identifiers", where for each we
want to create some data structure and add that to some general storage. An example
of such storage is shown below, using the \texttt{scoped\_enum} concept proposed
earlier:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
template<std::scoped_enum E, typename T>
class EnumMap {
public:
    bool has_value(E e) { return mEnumMap.contains(e); }
    void add_value(E e, T&& t) { mEnumMap[e] = t; }
    std::optional<T> get_value(E e) {
        if (auto it = mEnumMap.find(e); it != mEnumMap.end()) {
            return { it->second };
        } else { return std::nullopt; }
    }
private:
    std::map<E, T> mEnumMap;
};
\end{lstlisting}

We can then obtain an (possibly \textit{random-access}) iterator, which allows
us to iterate through all values in the enumeration and add them to the
storage:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
void fill(EnumMap<auto T, auto V>& em) {
    for (T t : std::enum_values<T>)
        em.add value(t, V{});
}
\end{lstlisting}

\noindent
\textbf{NOTE:} I have absolutely no idea how this implementation could be achieved,
but I strongly suspect that it requires one or more modifications to the standard,
and probably also extended parsing rules.


\subsection{Enum Cardinality}

Another feature which might sometimes be useful, is the ability to count the number
of values inside an enumeration. A real-world example, here found in the
\texttt{glslang} library for shader compilation:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
typedef enum {
    EShSourceNone,
    EShSourceGlsl,               // GLSL, includes ESSL (OpenGL ES GLSL)
    EShSourceHlsl,               // HLSL
    LAST_ELEMENT_MARKER(EShSourceCount),
} EShSource;                     // if EShLanguage were EShStage, this could be EShLanguage instead
\end{lstlisting}

Granted, this is a \texttt{C} library, but the use case for \texttt{C++} would be
identical. And it would likely be used for validating the input because \texttt{enums}
are unsafe (even though the code is still unsafe!):\vspace{2mm}

\begin{lstlisting}[language=Cpp]
void my_function(EShSource e) {
    if (e < 0 || e >= LAST_ELEMENT_MARKER) {
        // report an error...
    }
}
\end{lstlisting}


\noindent
This proposal then adds a new template meta-function, which at compile time computes
the number of elements - ie. the \textit{cardinality} - of a scoped enumueration:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
template<scoped_enum E>
struct enum_cardinality {
    static constexpr std::size_t value = 1; // <-- compiler implements this
};

template<scoped_enum E>
std::size_t enum_cardinality_v = std::enum_cardinality<E>::value;

// Try it out:
enum class Weekday { monday, tuesday, /* etc. */ };
static_assert(std::enum_cardinality_v<Weekday> == 7);
\end{lstlisting}

\noindent
\textbf{NOTE:} An argument can be made for extending \texttt{enum\_cardinality}
to also work for \textit{unscoped} enums, which at the very least could be used
to then improve the error check from the \texttt{glslang} code if it was placed
in a C++ context. This would be an alternative to simply replacing
\texttt{EShSource} with a scoped enum, which with this proposal in mind would
completely remove the need for the bounds error check.


\subsection{Enum position}

We might also want to know the position or the \textit{index} of an enum value
inside an enumeration. As with \texttt{std::enum\_cardinality}, we need to rely
on the compiler to provide an implementation, as well as a requirement of strict
compile-time evaluation:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
template<std::scoped_enum T, T E>
struct enum_position {
    static constexpr std::size_t value = 0; // <-- compiler implements this
};

template<std::scoped_enum T, T E>
std::size_t enum_position_v = std::enum_position<T, E>::value;

// Try it out:
enum class Fruit { banana = 1, apple = 2, orange = 4, clementine = 8 };
static assert(std::enum_value_v<Fruit::orange> == 4);
static assert(std::enum_position<Fruit, Fruit::orange> == 2);
\end{lstlisting}

\noindent
With both \texttt{enum\_cardinality} and \texttt{enum\_position} implemented, we
can create e.g. create a priority mapper with a priority for each enum value:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
template<std::scoped_enum T>
class EnumPriorityMapper {
public:
    unsigned int GetPriority(T t) {
        return mPriorities[std::enum_position_v<T, t>];
    }
    void SetPriority(T t, unsigned int p) {
        mPriorities[std::enum_position_v<T, t>] = p;
    }
private:
    unsigned int mPriorities[std::enum_cardinality_v<T>];
};
\end{lstlisting}


\subsection{Weak ordering}

With a good comparison operator in place, it becomes straight-forward to test
the order in which values appear inside an enumeration. Using the "spaceship operator"
introduced with C++20, this proposal defines two three-way comparison functions:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
template<std::scoped_enum T>
constexpr std::weak_ordering operator<=>(T t, std::integral auto i) {
    return std::enum_value(t) <=> i;
}

template<std::scoped_enum T>
std::weak_ordering operator<=>(T s, T t) {
    return std::enum_value(s) <=> std::enum_value(t);
}

// Try it out:
enum class MyScopedEnum : int {
    value_0 = 0, value_1 = 1, value_2 = 2, value_3 = 3, value_43 = 43, value_57 = 57
};
static_assert((MyScopedEnum::value_0 <=> 0) == 0);
static_assert((1 <=> MyScopedEnum::value_2) < 0);
static_assert((MyScopedEnum::value_2 <=> 1) > 0);
\end{lstlisting}


\subsection{Enum as bitmask (potentially has a new language feature)}

Bitmasking enums is a larger topic, one that has always been kind of awkward.
We can easily use C-style enums and directly mask them because they are integers,
but this approach provides no safety guarantees. The following example shows such
an awkward example:

\beforecodespace\begin{lstlisting}[language=Cpp]
enum MyUnscopedEnum { val_0 = 0, val_1 = 1, val_4 = 4 };
enum AnotherUnscopedEnum { val_5 = 5, val_6 = 6 };

// C++20 deprecates this, but for earlier versions (<= 17) it is valid.
int m = val_0 | val_5;
\end{lstlisting}

\noindent
At least, this is the output from GCC (I have not tested with other compilers!).
But we can improve upon the type safety a bit by using a scoped enum instead.
But now we have another awkward problem - we must statically cast to underlying
type before using the bitwise operator:

\beforecodespace\begin{lstlisting}[language=Cpp]
enum class MyScopedEnum { val_0 = 0, val_1 = 1, val_4 = 4 };
enum class AnotherScopedEnum { val_5 = 5, val_6 = 6 };

int p = MyScopedEnum::val_0 | AnotherScopedEnum::val_5; // error: no match for 'operator|'
int q = static_cast<int>(MyScopedEnum::val_0) | static_cast<int>(AnotherScopedEnum::val_5); // valid
auto q = std::enum_value(MyScopedEnum::val_0) | std::enum_value(AnotherScopedEnum::val_5); // same
\end{lstlisting}

We could define an \texttt{operator|} as suggested by the compiler, but then we are
left with another very awkward problem:

\beforecodespace\begin{lstlisting}[language=Cpp]
template<std::scoped_enum T>
constexpr ?? operator| (T t1, T t2) {
    std::enum_type_t<T> result = std::enum_value<T>(t1) | std::enum_value<T>(t2);
    return ??
}
\end{lstlisting}

\noindent
Now we don't know which type to return - there are two options, and both of them are
bad. If we return the underlying (integer) type then we bypass type safety, and if
we return the same enum type then we return a value that should not exist:

\beforecodespace\begin{lstlisting}[language=Cpp]
enum class MyFlags : int { val_1 = 0x01, val_2 = 0x02, val_4 = 0x04, val_8 = 0x08 };

MyFlags f1 = MyFlags::val_1 | MyFlags::val_2; // This is not a valid 'MyFlags' !
int f2 = MyFlags::val_1 | MyFlags::val_2 | MyFlags::val_4; // Problematic!
\end{lstlisting}

The second case in particular is problematic, because it bypasses type safety.
If the operator returns an integer, then we must define another operator which
takes as arguments a scoped enum and an integer, and that enables us to pass to
it any integer, like $-1$!

\noindent
Both choices yield bad code, or at the very least \textit{unsafe} code, and it
would be far better if the process of bitmasking enums could be automated in a
way that is type safe, practical, without or with limited runtime overhead, and
adheres to the general design philosophies of the standard library. It would
also be very nice if the operators supported \texttt{constexpr}. The first
solution that comes to mind is to create a class. The solution might not be
perfect, and likely needs some adjusting, but it does solve the problems
described above while being relatively easy to use:

\beforecodespace\begin{lstlisting}[language=Cpp]
template<typename T>
concept enum_valid_bitmask_type = ... // type should be restricted

template<typename T>
requires enum_valid_bitmask_type<T>
class enum_bitmask {
    using type = std::enum_type_t<T>;

public:
    constexpr bool empty();
    void add_flag(T _t);
    constexpr bool has_flag(T _t);
    void clear();
    // operators and constructors omitted here

private:
    type t {0U};
};
\end{lstlisting}

This class may be implemented in roughly 50 lines of code, plus some additional
operators for parameter switching. Below is a short code snippet (which has been
tested) that showcases some example usage:

\beforecodespace\begin{lstlisting}[language=Cpp]
enum class MyBitflag : unsigned int {
    val_1 = 0x01U, val_2 = 0x02U, val_4 = 0x04U, val_8 = 0x08U
};

constexpr std::enum_bitmask<MyBitflag> test1;
static_assert(test1.empty());

constexpr std::enum_bitmask<MyBitflag> test2 =
    MyBitflag::val_1 | MyBitflag::val_2 | MyBitflag::val_4 | MyBitflag::val_8;
static_assert(test2.has_flag(MyBitflag::val_4));
\end{lstlisting}


\noindent
The examples are hopefully enough to express the intent of the class.
The full implementation is included in section~\ref{proposed-notation-summary}.
The way the class and its operators are designed allow for \textit{compile-time}
checks through \texttt{static\_cast}, and lets us use scoped enums as fully type-safe
bitmasking operands.

The four public member functions
provide an interface that is easy to read and use, and by using a template we omit
the penalties of vtable lookups. The templated type which must be a scoped enum is
verified by a "concept", because additional constraints should be placed upon usage
of the class. One such constraint would be to check that the scoped enum used does
not have a value that is zero (\underline{language feature}):

\beforecodespace\begin{lstlisting}[language=Cpp]
template<scoped_enum E>
struct enum_all_nonzero
{
    static constexpr bool value = true; // <-- compiler implements this
};

template<scoped_enum E>
bool enum_all_nonzero_v = enum_all_nonzero<E>::value;
\end{lstlisting}

\noindent
\textbf{NOTE:} Another constraint might be to require that the underlying type
of the enum is unsigned, unless that would yield some platform-specific
problems (I'm not sure!). Also, C is notorious for its unsafe handing of 64-bit
bit masking of enums, so it might also be desireable to restrict to maximum
32-bit integers or whatever the platform uses for \texttt{`int`}. I don't
\textit{think} C++ has this problem, but I'm not completely sure.


\section{Proposed Notation Summary}
\label{proposed-notation-summary}

In summary, there's a lot of new features that would \textit{significantly} improve
the usage of scoped enums. Some of these may be implemented as library features,
while others simply cannot be written with current C++. Since most of the symbols
defined are dependent upon \texttt{std::is\_scoped\_enum}, C++23 is the
\textit{minimal} requirement for an implementation of this proposal. A quick overview
of the function and meta-function signatures are listed here:\vspace{2mm}

\begin{lstlisting}[language=Cpp]
// header: enum
// #include <enum>
#include <type_traits>

namespace std {
    // ============================ //
    // === CONCEPT: scoped_enum === //
    // ============================ //
    template<typename T>
    concept scoped_enum = std::is_scoped_enum_v<T>;

    // ================================= //
    // === VALUE: enum_cardinality_v === //
    // ================================= //
    template<scoped_enum E>
    struct enum_cardinality
    {
        static constexpr std::size_t value = 0; // <-- compiler implements this
    };

    template<scoped_enum E>
    std::size_t enum_cardinality_v = std::enum_cardinality<E>::value;

    // ============================== //
    // === VALUE: enum_position_v === //
    // ============================== //
    template<std::scoped_enum T, T E>
    struct enum_position {
        // TODO: Perhaps we _can_ actually use some template meta-programming to do this?
        static constexpr std::size_t value = 0; // <-- compiler implements this
    };

    template<std::scoped_enum T, T E>
    std::size_t enum_position_v = std::enum_position<T, E>::value;

    // ========================= //
    // === TYPE: enum_type_t === //
    // ========================= //
    template<scoped_enum E>
    struct enum_type
    {
        using type = std::underlying_type<E>::type;
    };

    // NOTE: Since C++23 we have std::to_underlying :
    // template<class Enum>
    // constexpr std::underlying_type_t<Enum> to_underlying(Enum e);
    // This is equivalent to:
    // return static_cast<std::underlying_type_t<Enum>>(e);

    template<scoped_enum E>
    using enum_type_t = std::enum_type<E>::type;

    // ========================= //
    // === VALUE: enum_value === //
    // ========================= //
    template<scoped_enum Enum>
    constexpr std::enum_type_t<Enum> enum_value(Enum e) noexcept
    {
        return static_cast<std::enum_type_t<Enum>>(e);
    }

    // ================================ //
    // === VALUE: enum_all_nonzero_v === //
    // ================================ //
    template<scoped_enum E>
    struct enum_all_nonzero
    {
        static constexpr bool value = true; // <-- compiler implements this
    };

    template<scoped_enum E>
    bool enum_all_nonzero_v = enum_all_nonzero<E>::value;

    // =========================== //
    // === CLASS: enum_bitmask === //
    // =========================== //
    template<typename T>
    concept enum_valid_bitmask_type =
        std::scoped_enum<T>
        && std::is_unsigned_v<std::enum_type_t<T>>
        && std::scoped_enum_no_zero<T>; // possible definition

    template<typename T>
    requires enum_bitmask_type<T>
    class enum_bitmask
    {
        using type = std::enum_type_t<T>;
    public:
        inline constexpr enum_bitmask() { t = type{0U}; }
        inline constexpr enum_bitmask(T _t) { t = std::enum_value<T>(_t); }
        inline constexpr enum_bitmask(enum_bitmask& b) { t = b.t; }
        inline enum_bitmask& operator= (T _t)
        {
            t = std::enum_value<T>(_t); return *this;
        }

        inline constexpr bool empty() const { return t == type{0U}; }
        inline void add_flag(T _t) { t |= std::enum_value<T>(_t); }
        inline constexpr bool has_flag(T _t) const { return t & std::enum_value<T>(_t); }
        inline void clear() { t = type{0U}; }

        inline constexpr bool operator& (T _t) const
        {
            return t & std::enum_value<T>(_t);
        }

        inline constexpr enum_bitmask operator| (T _t) const
        {
            return static_cast<T>(t | std::enum_value<T>(_t));
        }

        inline enum_bitmask& operator|=(T _t)
        {
            t |= std::enum_value<T>(_t); return *this;
        }
        inline constexpr type operator()() const
        {
            return t;
        }

        inline bool operator== (const enum_bitmask& b) { return t == b.t; }

    private:
        std::enum_type_t<T> t {0U};
    };
} // namespace std

// =============================== //
// === COMPARISON: operator<=> === //
// =============================== //
template<std::scoped_enum T>
constexpr std::weak_ordering operator<=>(T t, std::integral auto i)
{
    return std::enum_value(t) <=> i;
}

template<std::scoped_enum T>
std::weak_ordering operator<=>(T s, T t)
{
    return std::enum_value(s) <=> std::enum_value(t);
}

// ========================== //
// === BITFLAG operations === //
// ========================== //
template<std::scoped_enum T>
constexpr std::enum_bitmask<T> operator| (T s, T t)
{
    return std::enum_bitmask(s) | t;
}

template<std::scoped_enum T>
constexpr bool operator& (T _t, const std::enum_bitmask<T>& b)
{
    return b() & std::enum_value<T>(_t);
}

template<std::scoped_enum T>
constexpr bool operator& (std::enum_type_t<T> f, T t)
{
    return f & std::enum_value(t);
}

template<std::scoped_enum T>
constexpr bool operator& (T t, std::enum_type_t<T> f)
{
    return std::enum_value(t) & f;
}

\end{lstlisting}



\section{Design Decisions}

The overall design goals have been twofold:

\begin{enumerate}
\item Provide enhancements that all operate \underline{compile-time}.
\item Only provide enhancements for the type-safe scoped enums. Do nothing for the "old" C-style enums.
\end{enumerate}

\noindent
As such, most features have been written with the proposed syntax of template
meta-functions, with the use of modern C++ features such as \texttt{concepts} and
"three-way comparison" (\texttt{<=>}). The proposed syntax are merely
\textit{suggestions}, ie. "what I want", and not necessarily "how I want it".


\section{Technical Specifications}

I don't really know what to write here, since a complete code listing is shown
in the previous section.


\section{Acknowledgements}

Thanks to \textbf{Francesco Pretto} for suggesting the feature of bitmasking enums.
I took the advice and added it to this proposal following the initial post.

\noindent
Thanks to \textbf{Maxime Coste} for commenting on the "reflection proposal" probably
being the current direction for fixing the issues with enums which inspired this writing.
Since I'm unsure what the reflection is going to fix or not, I decided to write this
proposal anyways. Likely, and indeed hopefully, the reflection may be used to
implement some of the features suggested here.

\noindent
Thanks to \textbf{Jason McKesson} for suggesting consteval'd fixed-sized arrays created
by the compiler instead of defining an enum iterator and the enum to string functionality
suggested here. I agree that it would be simpler, but I also think it would end up
resulting in many programmers defining their own constructs instead of them being
in the standard library. My opinion may be flawed, but I appreciate the feedback
nonetheless.


%% \section{References}

%% \bibliographystyle{ACM-Reference-Format}
%% \bibliography{references}

\end{document}
